Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-26 main.tex
Finished an editing pass of the whole paper.

\section{Discussion}
%
We have shown that at a 6~\si{\kph} riding speed the addition of balance assist
control reduces the chance that a rider will fall when perturbed around the
limits of their control authority. But this effect diminishes just below
significance at the higher speed scenario of 10~\si{\kph}. We were only able to
test these two speed-gain scenarios for mostly homogeneous sets of riders
within the resources of this research project, but additional experimental work
could help understand more completely the range and limits of the positive
effect of the balance system. For example, it is possible that simply
increasing the controller gain at 10~\si{\kph} also results in a significant
positive effect. A longitudinal study of normal use of the balance assist
bicycle compared to a control group could provide the strongest evidence of any
benefit we have seen in this more narrow scenario.

@@

The controller design we utilize,
Equation~\ref{eq:implemented-controller}, also increases the weave mode
frequency by a factor of about three up to about 1~\si{\hertz}. This bandwidth
is still controllable by the human's neuromuscular system, but may feel
unnatural as it is more akin to what the steering would feel like at in the
\SIrange{30}{40}{\kph} speed range. \Citet{Hanakam2023} reported
dissatisfaction in subjective rider feeling on their similar bicycle to ours
and this effect to the human-controlled plant dynamics could be connected to
this.

\subsection{Application of the Logistic Model}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance assist system
reduces the odds to 500:1.
However, that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse the rider was subjected to.
This means that the balance assist is most effective for perturbations close to
the participant's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

To illustrate the effect of the balance assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use Equation~\ref{eq:log-regress} and the coefficients
in \ref{tab:freq-coefs-6}. For simplicity's sake, the interaction effects are
not included. For example, assuming that the mean angular impulse $\bar{L}$ of all the
perturbations applied to a participant is 100~\si{\newton}, and the standard
deviation $\sigma^{L}=15$. The centered and scaled angular impulse can be
calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
can be calculated as follows.
%
\begin{equation}
  \begin{split}
    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
    = & \quad
    \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}} \\
    = &
   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot c \\
    = &
    1.42 -0.64 \cdot c
  \end{split}
\end{equation}
%
The state of the balance assist \(s\) is a binary variable. If the
balance assist is turned on, the log-odds that a fall occurs are decreased by
0.64. The odds and probability can be calculated:
%
\begin{align}
  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64c} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53c \\
  p_{ij}|_{c=0} & = \frac{4.14}{1 + 4.14} = 0.81 \\
  p_{ij}|_{c=1} & = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}} = 0.69
\end{align}
%
Turning on the balance assist system reduces the probability that the
perturbation results in a fall from 81\% to 69\%.

@@

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a statistically
significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
same perturbation magnitude causes larger lateral deviations. At 10~\si{\kph}
almost all falls were due to the bicycle exiting the maximum width of the
treadmill. If the same experiment was performed on an infinitely wide plane,
the riders may have recovered from more perturbations. At 6~\si{\kph} the
riders could often recover in the allotted treadmill width due to the smaller
lateral deviations. We believe our results are very much dependent on the two
modes of falling with use: exit the treadmill width or foot is placed on the
belt. Cycle paths are a similar width as the treadmill, so rider's are often
limited in width when recovering from a fall.

\subsection{Learning Effect}
%
An increase in the number of perturbation that a participant already
experienced, decreases the probability that a fall occurs. This is likely due
to the participants learning how to better recover from the perturbation over
the duration of the experiment. This effect is significant at both 6 and
10~\si{\kph}, and in the same order of magnitude. This suggests that the
learning effect that occurs during the experiment is not strongly dependent on
the speed.

@@

Roll and steer angle are not a significant predictor of fall probability at 6
or at 10~\si{\kph}. We expected these variables to have a significant effect
based on the following reasoning. If you are in a rolled and steered state that
is far from the upright equilibrium, then a perturbation that further pushes
you from the equilibrium should have some additive or multiplicative effect on
the resulting motion trajectory.

@@

\subsection{Extrapolation to Natural Falls}
%
The positive effect of the balance assist system is coupled to the assumptions
and experimental scenarios we implemented and there is unfortunately no simple
way to extrapolate our results to reductions of single-actor crashes we may see
if such a system were deployed widely to bicyclists. Although, our results do
indicate that we would see such a reduction, even if only in a class of
single-actor crashes that most resemble our experimental design. If there were
more comprehensive and detailed natural data of how people fall we could make
estimates on the number of falls reduced if everyone road a balance assist
bicycle.

\section{Conclusion}
%
Automatically controlling a steering motor in a bicycle using roll rate
feedback lowers the speed at which the bicycle is stable. This makes the
bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
which is easier for a rider to balance. Perturbation forces applied to the
handlebar can cause a rider to fall and every rider has a threshold force at
which they are more likely to fall than not. The probability of falling when
mechanically perturbed is significantly reduced when traveling at 6~\si{\kph}
when the balance assisting control is activated. This effect is present when
traveling at 10~\si{\kph} but more investigation is needed to determine if the
effect can be significant. The positive effect to balance is rider independent
and most effective in the regime of perturbations near the rider's control
authority threshold. Given that similar effects cause falls during bicycling,
use of the balance assist system in real world use cases will reduce the number
of falls at low bicycling speeds.

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle. We acknowledge Shannon van de Velde's and Marco Reijne's
development of the bicycle perturbation system. We thank all of the assistants
that helped run the experiments and the participants for their time an enthusiasm.
We thank Peter Stahlecker for valuable feedback in early revisions.

>>>>

\section{Discussion}
%
We have shown that at a 6~\si{\kph} riding speed the addition of balance assist
control reduces the chance that a rider will fall when perturbed around the
limits of their control authority. But this effect diminishes just below
significance at the higher speed scenario of 10~\si{\kph}. We were only able to
test these two speed-gain scenarios for mostly homogeneous sets of riders
within the resources of this research project, but additional experimental work
could help understand more completely the range and limits of the positive
effect of the balance system. For example, it is possible that simply
increasing the controller gain at 10~\si{\kph} also results in a significant
positive effect. A longitudinal study of normal use of the balance assist
bicycle compared to a control group could provide the strongest evidence of any
benefit we have seen in this specific scenario.

@@

The controller design we utilize, Equation~\ref{eq:implemented-controller},
also increases the weave mode frequency by a factor of about three up in a
bandwidth of about 1~\si{\hertz}. This bandwidth is still controllable by the
human's neuromuscular system, but may feel unusual as it is more akin to what
the steering would feel like at in the \SIrange{30}{40}{\kph} speed range.
\Citet{Hanakam2023} reported dissatisfaction in subjective rider feeling on
their similar bicycle to ours and this effect to the human-controlled plant
dynamics could be an explanation to their findings.

\subsection{Utility of the Logistic Model and Extrapolation to Natural Falls}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
insight.
To interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance assist system
reduces the odds to 500:1.
However, in that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall occurs are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse the rider was subjected to.
This means that the balance assist is most effective for perturbations close to
the participant's perturbation resistance threshold and that large
perturbations will make you fall regardless of the balance assist's help.

To illustrate the effect of the balance assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use Equation~\ref{eq:log-regress} and the
coefficients in Table~\ref{tab:freq-coefs-6}. For simplicity's sake, the
interaction effects are not included. For example, assuming that the mean
angular impulse $\bar{L}$ of all the perturbations applied to a participant is
100~\si{\newton} and the standard deviation $\sigma_{L}=15$. The centered and
scaled angular impulse can be calculated by subtracting $\bar{L}$ from the
applied angular impulse $L$, and dividing this by $\sigma_{L}$. The same
applies for the perturbation order $j$, initial roll angle $\phi_0$, and
initial steer angle $\delta_0$. If we take the coefficients estimated for
cycling at 6~\si{\kph}, the log-odds of falling can be calculated as follows.
\todo{Check this equation, where is \(\bar{L}\) 100N in the equation?}
%
\begin{equation}
  \begin{split}
    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
    = & \quad
    \beta + \sum_{k=1}^{k=5}\alpha_k \frac{x_{ij}^{k}-\bar{x}_{ij}^{k}}{\sigma^{x^{k}}} \\
    = &
   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot c \\
    = &
    1.42 -0.64 \cdot c
  \end{split}
\end{equation}
%
The state of the balance assist \(c\) is a binary variable. If the
balance assist is turned on, the log-odds that a fall occurs are decreased by
0.64. The odds and probability can be calculated:
%
\begin{align}
  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64c} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53c \\
  p_{ij}|_{c=0} & = \frac{4.14}{1 + 4.14} = 0.81 \\
  p_{ij}|_{c=1} & = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}} = 0.69
\end{align}
%
Turning on the balance assist system reduces the probability that the
perturbation results in a fall from 81\% to 69\% in this case.

@@

The positive effect of the balance assist system is coupled to the assumptions
and experimental scenarios we implemented and there is unfortunately no simple
way to extrapolate our results to reductions of single-actor crashes we may see
if such a system were deployed widely to bicyclists. Although, our results do
indicate that we would see such a reduction, even if only in a class of
single-actor crashes that most resemble our experimental design. If there were
more comprehensive and detailed natural data of how people fall we could make
estimates on the number of falls reduced if everyone road a balance assist
bicycle.

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a
statistically significant effect at 10~\si{\kph}. As a bicycle travels at
higher speeds, the same perturbation magnitude causes larger lateral
deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
the maximum width of the treadmill. If the same experiment was performed on an
infinitely wide plane, the riders may have recovered from more perturbations.
At 6~\si{\kph} the riders could often recover in the allotted treadmill width
due to the smaller lateral deviations. We believe our results are very much
dependent on the two modes of falling, i.e. exit the treadmill width or foot is
placed on the belt. On the other hand, Cycle paths are a similar width as the
treadmill, so rider's are often limited in width when recovering from a fall
thus exiting the treadmill width may be an appropriate measure for indicating a
fall.

\subsection{Learning Effect}
%
An increase in the number of perturbations that a participant already
experienced, decreases the probability that a fall occurs. This is likely due
to the participants learning how to better recover from the perturbation over
the duration of the experiment. This effect is significant at both 6 and
10~\si{\kph}, and in the same order of magnitude. This suggests that the
learning effect that occurs during the experiment is not strongly dependent on
the speed.

@@

Roll and steer angle at the initiation of a perturbation are not a significant
predictor of fall probability at 6 or at 10~\si{\kph}. We expected these
variables to have a significant effect based on the following reasoning. If you
are in a rolled and steered state that is far from the upright equilibrium,
then a perturbation that further pushes you from the equilibrium should have
some additive or multiplicative effect on the resulting motion trajectory and
make it harder to recover from a fall.

@@

\section{Conclusion}
%
Automatically controlling a steering motor in a bicycle using roll rate
feedback lowers the speed at which the bicycle is stable. This makes the
bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
which is easier for a rider to control in balance. Perturbation forces applied
to the handlebar can cause a rider to fall and every rider has a perturbation
magnitude threshold at which they are more likely to fall than not. The
probability of falling when mechanically perturbed is significantly reduced
when traveling at 6~\si{\kph} when the balance assisting control is activated.
This effect is present when traveling at 10~\si{\kph} but more investigation is
needed to determine if the effect can be significant. The positive effect to
balance is rider independent and most effective in the regime of perturbations
near the limits of the rider's control authority. Given that similar effects
cause falls during bicycling, use of the balance assist system in real world
use cases will reduce the number of falls at low bicycling speeds.

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle. We acknowledge Shannon van de Velde and Marco Reijne's
development of the bicycle perturbation system. We thank all of the assistants
that helped run the experiments and the participants for their time an
enthusiasm. We thank Peter Stahlecker for valuable feedback in early revisions.
